---
title: "Heatmap"
output: html_document
date: "2025-07-28"
---

# Clean and Rarefy the Phyloseq file
```{r}
#Load Package
rm(list = ls())
gc()
suppressMessages({
  library("tidyverse")
  library("phyloseq")
  library("rstatix")
  library("vegan")
  library("picante")
  library("ggpubr")
  library(vegan)
  library(dplyr)
  library(tidyr)
  library(ggplot2)  
  library(tibble)
  library(poppr)  # For AMOVA if genetic analysis is needed
  library(viridis)
  library(patchwork)
  library(pheatmap)
  library(RColorBrewer)
  library(agricolae)
  library(microbiomeSeq)
  library(microbiomeutilities)
  library(viridis)
})
no_of_cores = 16
setwd("/work/benson/bpeng4/Favor/Favor_16S")
load("intermediate/Favor_Forward_ps.rda")
ls()
ps

###Pre-arrangement on  the phyloseq file
################
ps.clean <- subset_taxa(ps, Kingdom == "Bacteria") %>%
  subset_taxa(!is.na(Phylum)) %>%
  subset_taxa(!Class %in% c("Chloroplast")) %>%
  subset_taxa(!Family %in% c("Mitochondria"))
ps.clean

## Standardize Data by Rarefication
##
#Check Total Reads to find the Thresholds for Rarefication
OTU<-otu_table(ps.clean) |>as.data.frame()
OTU$Total<- rowSums(OTU)
#Rarefication
ps.rare<-rarefy_even_depth(ps.clean, sample.size = 13634,
                           rngseed = 111, replace = TRUE, trimOTUs = TRUE, verbose = TRUE)

```


#glommate to Genus Level
ps.genus = phyloseq::tax_glom(ps.rare, taxrank = rank_names(ps.rare)[6])

# Adjust by FBB0
```{r}
normalize_by_fbb0 <- function(ps.genus.S, Subject) {
  # Step 1: Extract OTU table (samples are rows, taxa are columns)
  otu <- as(otu_table(ps.genus.S), "matrix")
  
  # Step 2: Extract FBB0 replicates
  fbb0_1 <- otu[paste0(Subject, "-FBB0-1"), ]
  fbb0_2 <- otu[paste0(Subject, "-FBB0-2"), ]
  
  # Step 3: Extract sample metadata
  sample_df <- as(sample_data(ps.genus.S), "data.frame")
  
  # Step 4: Normalize each sample by its matching FBB0 replicate
  otu_norm <- otu
  for (sample_id in rownames(otu)) {
    rep_val <- sample_df[sample_id, "Rep"]
    if (rep_val == 1) {
      otu_norm[sample_id, ] <- (otu[sample_id, ] + 0.01) / (fbb0_1 + 0.01)
    } else if (rep_val == 2) {
      otu_norm[sample_id, ] <- (otu[sample_id, ] + 0.01) / (fbb0_2 + 0.01)
    }
  }
  # Step 5: Remove taxa (columns) with any NaN, Inf, or -Inf
  bad_taxa <- apply(otu_norm, 2, function(col) any(!is.finite(col)))
  otu_norm_clean <- otu_norm[, !bad_taxa]
  
  # Step 6: Remove FBB0 samples from both OTU matrix and sample data
  fbb0_samples <- rownames(sample_df)[grepl("FBB0", rownames(sample_df))]
  otu_final <- otu_norm_clean[!rownames(otu_norm_clean) %in% fbb0_samples, ]
  sample_final <- sample_df[!rownames(sample_df) %in% fbb0_samples, ]
  
  # Step 7: Rebuild phyloseq object with cleaned data
  tax_final <- tax_table(ps.genus.S)[colnames(otu_final), ]
  otu_phyloseq <- otu_table(otu_final, taxa_are_rows = FALSE)
  sample_phyloseq <- sample_data(sample_final)
  ps.cleaned <- phyloseq(otu_phyloseq, tax_final, sample_phyloseq)
  
  return(ps.cleaned)
}

```


# Write Permanova Calculation function
```{r}
ps.filter <- function(ps.genus) {
  # Extract OTU/ASV table and metadata
  otu_table_df <- as.data.frame(otu_table(ps.genus))
  metadata <- data.frame(sample_data(ps.genus))
  taxa_table <- as.data.frame(tax_table(ps.genus))

  # Ensure 'Sample' is a factor
  metadata$Sample <- as.factor(metadata$Sample)

  # PERMANOVA
  adonis_results <- adonis2(otu_table_df ~ Sample, data = metadata, permutations = 999, method = "bray")
  p_value <- adonis_results$`Pr(>F)`[1]

  # Individual ANOVA per taxon
  p_values <- apply(otu_table_df, 2, function(x) {
    df <- data.frame(x = x, Sample = metadata$Sample)
    fit <- aov(x ~ Sample, data = df)
    summary(fit)[[1]][["Pr(>F)"]][1]
  })

  # Adjust p-values using FDR
  p_values_adj <- p.adjust(p_values, method = "fdr")

  # Get significant taxa names
  signif_taxa <- names(p_values_adj[p_values_adj < 0.05])

  # Check if any significant taxa remain
  if (length(signif_taxa) == 0) {
    message("No significant taxa found (adjusted p < 0.05). Returning NULL.")
    return(NULL)
  }

  # Retain only taxa that actually exist in the phyloseq object
  existing_taxa <- taxa_names(ps.genus)
  signif_taxa_filtered <- signif_taxa[signif_taxa %in% existing_taxa]

  # If the intersection is empty, return NULL
  if (length(signif_taxa_filtered) == 0) {
    message("Significant taxa do not overlap with current phyloseq object. Returning NULL.")
    return(NULL)
  }

  # Prune phyloseq object
  ps.genus.filtered <- prune_taxa(signif_taxa_filtered, ps.genus)

  # Final check: ensure non-zero dimensions
  if (ntaxa(ps.genus.filtered) == 0 || nsamples(ps.genus.filtered) == 0) {
    message("Filtered phyloseq object has 0 taxa or samples. Returning NULL.")
    return(NULL)
  }

  return(ps.genus.filtered)
}


```


# Write heatmap plotting function
```{r}
heatmapsample <- function(input_ps){

  # Desired sample order
  desired_order <- c("HP301","PHG35","NC262","764","A554","LH60","PHG39","B57","PHG47"    
                        )
  
  # Extract OTU table
  otu_df <- as.data.frame(otu_table(input_ps))
  if (taxa_are_rows(input_ps)) {
    otu_df <- t(otu_df)
  }
  
  # Use tax_table to label taxa uniquely
  tax_df <- as.data.frame(tax_table(input_ps))
  tax_labels <- paste0(tax_df$Genus, "_", rownames(tax_df))
  colnames(otu_df) <- tax_labels
  
  # Get sample metadata (in case you need sample info)
  meta_df <- as.data.frame(sample_data(input_ps))
  
  # Add sample names
  otu_df$Sample <- meta_df$Sample
  
  # Aggregate by sample
  otu_avg <- otu_df %>%
    group_by(Sample) %>%
    summarise(across(where(is.numeric), mean, na.rm = TRUE))
  
  # Convert to matrix
  otu_avg_mat <- column_to_rownames(otu_avg, "Sample") %>% as.matrix()
  
  # Remove unwanted sample
  otu_avg_mat <- otu_avg_mat[rownames(otu_avg_mat) != "FBB0", , drop = FALSE]
  
  # Reorder
  otu_avg_mat <- otu_avg_mat[desired_order[desired_order %in% rownames(otu_avg_mat)], , drop = FALSE]
  
  # Log2 transform
  otu_transformed <- log2(otu_avg_mat + 1)
  
  # Plot
  pheatmap(
    mat = otu_transformed,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    color = colorRampPalette(rev(brewer.pal(7, "RdYlBu")))(100),
    show_colnames = TRUE,
    show_rownames = TRUE,
    main = "Heatmap (log2-transformed)"
  )
  return(pheatmap)
}
```

# Adjust by FBB0 for 19 Microbiomes
```{r}
Subjects <- c("S1", "S2", "S3", "S4", "S5", "S6", "S7", 
              "S9", "S12", "S14", "S15", "S16",
              "S17", "S18", "S19", "S20", "S21", "S22", "S24")

# Create empty list to store results
ps.cleaned.list <- list()

# Loop through each subject and normalize
for (Subject in Subjects) {
  # Subset the phyloseq object for this Subject
  ps.sub <- subset_samples(ps.genus, Microbiome == Subject)
  
  # Normalize by FBB0 and clean
  ps.cleaned <- normalize_by_fbb0(ps.sub, Subject)
  
  # Save result into list
  ps.cleaned.list[[Subject]] <- ps.cleaned
}
```

# Do Permanova for Each Individual Microbiome
```{r}
# Initialize result list
ps.filtered.list <- list()

# Loop through subjects
for (Subject in Subjects) {
  ps.filtered <- ps.filter(ps.cleaned.list[[Subject]])
  
  # Only store if filtering was successful (not NULL)
  if (!is.null(ps.filtered)) {
    ps.filtered.list[[Subject]] <- ps.filtered
  } else {
    message(paste("Skipping", Subject, "- no significant taxa."))
  }
}

```


lefse_result <- run_lefse(
  ps = ps.genus,
  group = "Sample",            
  norm = "CPM",
  lda_cutoff = 2.0,
  kw_cutoff = 0.05,
  wilcoxon_cutoff = 0.05
)


# LDA score barplot: use ggplot2 + marker_table
lda_tbl <- marker_table(lefse_result)


# Cladogram plot
plot_cladogram(lefse_result, color = c('#F781BF', '#CCEBC5', '#FFFF99', '#FF7F00', '#8DD3C7', '#7FC97F', '#999999',  
  '#CAB2D6'))
























ps.genus.S1 <- subset_samples(ps.genus, Microbiome == "S1")
ps.genus.filtered.S1 <- ps.filter(ps.genus.S1)
ps.genus.S2 <- subset_samples(ps.genus, Microbiome == "S2")
ps.genus.filtered.S2 <- ps.filter(ps.genus.S2)
ps.genus.S3 <- subset_samples(ps.genus, Microbiome == "S3")
ps.genus.filtered.S3 <- ps.filter(ps.genus.S3)
ps.genus.S4 <- subset_samples(ps.genus, Microbiome == "S4")
ps.genus.filtered.S4 <- ps.filter(ps.genus.S4)
ps.genus.S5 <- subset_samples(ps.genus, Microbiome == "S5")
ps.genus.filtered.S5 <- ps.filter(ps.genus.S5)
ps.genus.S6 <- subset_samples(ps.genus, Microbiome == "S6")
ps.genus.filtered.S6 <- ps.filter(ps.genus.S6)
ps.genus.S7 <- subset_samples(ps.genus, Microbiome == "S7")
ps.genus.filtered.S7 <- ps.filter(ps.genus.S7)
ps.genus.S9 <- subset_samples(ps.genus, Microbiome == "S9")
ps.genus.filtered.S9 <- ps.filter(ps.genus.S9)
ps.genus.S12 <- subset_samples(ps.genus, Microbiome == "S12")
ps.genus.filtered.S12 <- ps.filter(ps.genus.S12)
ps.genus.S14 <- subset_samples(ps.genus, Microbiome == "S14")
ps.genus.filtered.S14 <- ps.filter(ps.genus.S14)
ps.genus.S15 <- subset_samples(ps.genus, Microbiome == "S15")
ps.genus.filtered.S15 <- ps.filter(ps.genus.S15)
ps.genus.S16 <- subset_samples(ps.genus, Microbiome == "S16")
ps.genus.filtered.S16 <- ps.filter(ps.genus.S16)
ps.genus.S17 <- subset_samples(ps.genus, Microbiome == "S17")
ps.genus.filtered.S17 <- ps.filter(ps.genus.S17)
ps.genus.S18 <- subset_samples(ps.genus, Microbiome == "S18")
ps.genus.filtered.S18 <- ps.filter(ps.genus.S18)
ps.genus.S19 <- subset_samples(ps.genus, Microbiome == "S19")
ps.genus.filtered.S19 <- ps.filter(ps.genus.S19)
ps.genus.S20 <- subset_samples(ps.genus, Microbiome == "S20")
ps.genus.filtered.S20 <- ps.filter(ps.genus.S20)
ps.genus.S21 <- subset_samples(ps.genus, Microbiome == "S21")
ps.genus.filtered.S21 <- ps.filter(ps.genus.S21)
ps.genus.S22 <- subset_samples(ps.genus, Microbiome == "S22")
ps.genus.filtered.S22 <- ps.filter(ps.genus.S22)
ps.genus.S24 <- subset_samples(ps.genus, Microbiome == "S24")
ps.genus.filtered.S24 <- ps.filter(ps.genus.S24)

# Step 1: Create a named list of all taxa vectors
taxa_list <- list(
  S1 = taxa_names(ps.genus.filtered.S1),
  S2 = taxa_names(ps.genus.filtered.S2),
  S3 = taxa_names(ps.genus.filtered.S3),
  S4 = taxa_names(ps.genus.filtered.S4),
  S5 = taxa_names(ps.genus.filtered.S5),
  S6 = taxa_names(ps.genus.filtered.S6),
  S7 = taxa_names(ps.genus.filtered.S7),
  S9 = taxa_names(ps.genus.filtered.S9),
  S12 = taxa_names(ps.genus.filtered.S12),
  S14 = taxa_names(ps.genus.filtered.S14),
  S15 = taxa_names(ps.genus.filtered.S15),
  S16 = taxa_names(ps.genus.filtered.S16),
  S17 = taxa_names(ps.genus.filtered.S17),
  S18 = taxa_names(ps.genus.filtered.S18),
  S19 = taxa_names(ps.genus.filtered.S19),
  S20 = taxa_names(ps.genus.filtered.S20),
  S21 = taxa_names(ps.genus.filtered.S21),
  S22 = taxa_names(ps.genus.filtered.S22),
  S24 = taxa_names(ps.genus.filtered.S24)
)

# Step 2: Count in how many lists each taxon appears
taxa_counts <- table(unlist(taxa_list))

# Step 3: Keep taxa that appear in at least 4 or 5 or 9 microbiomes
common_taxa_4plus <- names(taxa_counts[taxa_counts >= 4])
common_taxa_5plus <- names(taxa_counts[taxa_counts >= 5])
common_taxa_10plus <- names(taxa_counts[taxa_counts >= 10])
common_taxa_14plus <- names(taxa_counts[taxa_counts >= 14])

# Plot for the common taxa
# Step 1: Put all phyloseq objects into a named list
ps_list <- list(
  S1 = ps.genus.filtered.S1,
  S2 = ps.genus.filtered.S2,
  S3 = ps.genus.filtered.S3,
  S4 = ps.genus.filtered.S4,
  S5 = ps.genus.filtered.S5,
  S6 = ps.genus.filtered.S6,
  S7 = ps.genus.filtered.S7,
  S9 = ps.genus.filtered.S9,
  S12 = ps.genus.filtered.S12,
  S14 = ps.genus.filtered.S14,
  S15 = ps.genus.filtered.S15,
  S16 = ps.genus.filtered.S16,
  S17 = ps.genus.filtered.S17,
  S18 = ps.genus.filtered.S18,
  S19 = ps.genus.filtered.S19,
  S20 = ps.genus.filtered.S20,
  S21 = ps.genus.filtered.S21,
  S22 = ps.genus.filtered.S22,
  S24 = ps.genus.filtered.S24
)

# Step 2: Apply pruning to all

common_taxa_5plus <- common_taxa_5plus[common_taxa_5plus != "ASV_1"]
common_taxa_10plus <- common_taxa_10plus[common_taxa_10plus != "ASV_1"]

ps_list_filtered <- lapply(ps_list, function(ps) prune_taxa(common_taxa_10plus, ps))


# remove NULLs (the empty ones)
ps_list_filtered <- Filter(Negate(is.null), ps_list_filtered)

# optional: check how many taxa remain in each object
sapply(ps_list_filtered, ntaxa)


# Heatmap for the common taxa in each microbiome
microbiomes <- c("S1", "S2", "S3", "S4","S5", "S6", "S7", 
                 "S9", "S12", "S14", "S15", "S16",
                 "S17", "S18", "S19", "S20", "S21", "S22", "S24")
            

                 
for (i in microbiomes){
  heatmapsample(ps_list_filtered[[i]])
}                 
                 







# Heatmap for Bifidobacterium, Bacteroides, Runimococcus, [Eubacterium] eligens, Fusicatenibacter, and Lachnospiraceae NK4A136  in S6, S9


ps.genus.S6<-subset_samples(ps.genus, ps.genus@sam_data$Microbiome == "S6")
ps.genus.S9<-subset_samples(ps.genus, ps.genus@sam_data$Microbiome == "S9")

common_taxa <- c("ASV_95","ASV_103","ASV_61","ASV_116")

ps.genus.filtered.S6 <- prune_taxa(common_taxa, ps.genus.S6)
ps.genus.filtered.S9 <- prune_taxa(common_taxa, ps.genus.S9)

heatmapsample(ps.genus.filtered.S6)
heatmapsample(ps.genus.filtered.S9)


# Two color heatmap
```{r}
heatmapsample <- function(input_ps){
  
  # Desired sample order
  desired_order <- c("A554", "PHG47", "PHG39", "B57", "HP301", 
                     "LH60", "764", "PHG35", "NC262")
  
  # Extract OTU table
  otu_df <- as.data.frame(otu_table(input_ps))
  if (taxa_are_rows(input_ps)) {
    otu_df <- t(otu_df)
  }
  
  # Use tax_table to label taxa uniquely
  tax_df <- as.data.frame(tax_table(input_ps))
  tax_labels <- paste0(tax_df$Genus, "_", rownames(tax_df))
  colnames(otu_df) <- tax_labels
  
  # Get sample metadata (in case you need sample info)
  meta_df <- as.data.frame(sample_data(input_ps))
  
  # Add sample names
  otu_df$Sample <- meta_df$Sample
  
  # Aggregate by sample
  otu_avg <- otu_df %>%
    group_by(Sample) %>%
    summarise(across(where(is.numeric), mean, na.rm = TRUE))
  
  # Convert to matrix
  otu_avg_mat <- column_to_rownames(otu_avg, "Sample") %>% as.matrix()
  
  # Remove unwanted sample
  otu_avg_mat <- otu_avg_mat[rownames(otu_avg_mat) != "FBB0", , drop = FALSE]
  
  # Reorder
  otu_avg_mat <- otu_avg_mat[desired_order[desired_order %in% rownames(otu_avg_mat)], , drop = FALSE]
  
  # Log2 transform
  otu_transformed <- log2(otu_avg_mat + 1)
  
  # Plot
  red_white <- colorRampPalette(c("white","red"))(100)
  
  pheatmap(
    mat = t(otu_transformed),
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    color = red_white,
    show_colnames = TRUE,
    show_rownames = TRUE,
    main = "Heatmap (Red–White Gradient)"
  )
  return(pheatmap)
}

```

